Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-04 ind_files/99_Appendix_D_examples.tex
Minor rephrasing

Moving to density space via the change of variables $x_i = n_i/K$, letting $\mathbf{x} = [x_1, x_2]^{\mathrm{T}}$, and comparing terms with equations \eqref{nD_functional_forms_for_replicator}, we can see that the per-capita fitness $w_i$ of each type is:
\begin{equation*}
	w_{1}(\mathbf{x}) = w_{2}(\mathbf{x}) = 1 - x_1 - x_2
\end{equation*}
The two types of birds have the same fitness! This implies that $w_1 = w_2 = \overline{w}$ and the selection term in \eqref{nD_replicator_mutator} vanishes.
However, if we now compute the per-capita turnover rates $\tau_i$  of each type, we see that we have
\begin{align*}
	\tau_{1}(\mathbf{x}) &= 1 + x_1 - x_2\\
	\tau_{2}(\mathbf{x}) &= 1 + x_1 + x_2
\end{align*}
Thus, $\tau_{1} < \tau_{2}$, and from equation \eqref{nD_stochastic_RM}, we thus know that noise-induced selection favors type 1 over type 2.
\myfig{1}{figures/App_stoch_LV_density_plots.png}{\textbf{(A)} Time series and \textbf{(B)} Density estimates for $p$, the fraction of type 1 individuals in the population for various values of $K$, obtained from a direct individual-based simulation of the model  conducted via the Gillespie algorithm. At high $K$, the population conforms to deterministic (infinite population) predictions. The time series are from single realizations. The density plots in panel $(B)$ are estimated from 100 independent realizations, each of which were run for $10^4$ timesteps. All simulations were initialized with $n_1 = n_2 = K/2$. In all simulations, $\mu = 0.05$.}{App_stoch_LV_dens_plots}
Direct Gillespie simulations indeed reveal that for low values of $K$, the fraction of individuals in the population that are of type 1 is significantly biased to be greater than 0.5, showing the effect of noise-induced selection (Figure \ref{App_stoch_LV_dens_plots}). This bias disappears for high values of $K$, as expected.

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that equation \eqref{nD_stochastic_RM} becomes
\begin{equation}
	dp =  \left[\frac{2}{K}p^2q + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right)\right]dt + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+}$ is as defined throughout this thesis and each $W^{(i)}_t$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. The second term is a mutational term, and simply reflects the fact that we assumed that $1 \to 2$ and $2 \to 1$ mutations are equally likely, and thus, the net effect of mutational effects depends on the difference between the frequencies of type 1 and type 2 individuals and flows towards the type with lower frequency, which is why it is $q-p$ and not $p-q$. Finally, the two $dW_t$ terms are non-directional and vanish upon taking an expectation over the probability space, and therefore have no net contribution other than `blurring out the results' if we look at the dynamics averaged over many realizations.

>>>>

Moving to density space via the change of variables $x_i = n_i/K$, letting $\mathbf{x} = [x_1, x_2]^{\mathrm{T}}$, and comparing terms with equations \eqref{nD_functional_forms_for_replicator}, we can see that the per-capita fitness $w_i$ of each type is:
\begin{equation*}
	w_{1}(\mathbf{x}) = w_{2}(\mathbf{x}) = 1 - x_1 - x_2
\end{equation*}
The two types of birds have the same fitness! This implies that $w_1 = w_2 = \overline{w}$ and the selection term in \eqref{nD_replicator_mutator} vanishes.
However, if we now compute the per-capita turnover rates $\tau_i$  of each type, we see that we have
\begin{align*}
	\tau_{1}(\mathbf{x}) &= 1 + x_1 - x_2\\
	\tau_{2}(\mathbf{x}) &= 1 + x_1 + x_2
\end{align*}
Thus, $\tau_{1} < \tau_{2}$, and from equation \eqref{nD_stochastic_RM}, we thus know that noise-induced selection favors type 1 over type 2.
\myfig{1}{figures/App_stoch_LV_density_plots.png}{\textbf{(A)} Time series and \textbf{(B)} Density estimates for $p$, the fraction of type 1 individuals in the population for various values of $K$, obtained from a direct individual-based simulation of the model defined by equations \eqref{App_example_stoch_LV_birth_rates} and \eqref{App_example_stoch_LV_death_rates}, simulated via the (exact) Gillespie algorithm. Dotted lines are at $p=0.5$. At high $K$, the population conforms to deterministic (infinite population) predictions, but at low K, the distribution is biased towards $p > 0.5$. The time series are from single realizations. The density plots in panel $(B)$ are estimated from 100 independent realizations, each of which were run for $10^4$ timesteps. All simulations were initialized with $n_1 = n_2 = K/2$. In all simulations, $\mu = 0.05$.}{App_stoch_LV_dens_plots}

Direct simulations of the individual-based model indeed reveal that for low values of $K$, the fraction of individuals in the population that are of type 1 is significantly biased to be greater than 0.5, showing the effect of noise-induced selection (Figure \ref{App_stoch_LV_dens_plots}). This bias disappears for high values of $K$, as expected.

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that equation \eqref{nD_stochastic_RM} becomes
\begin{equation}
	dp =  \left[\frac{2}{K}p^2q + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right)\right]dt + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+}$ is as defined throughout this thesis and each $W^{(i)}_t$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. The second term is a mutational term, and simply reflects the fact that we assumed that $\textrm{(type }1) \to \textrm{(type }2)$ and $\textrm{(type }2) \to \textrm{(type }1)$ mutations occur at the same rate $\mu$, and thus, the net effect of mutational effects depends on the difference between the frequencies of the two types and flows towards the type with lower frequency. Finally, the two $dW^{(i)}_t$ terms are non-directional and vanish upon taking an expectation over the probability space, and therefore have no net contribution other than `blurring out the results' if we look at the dynamics averaged over many realizations.
